---
title: "Final Project Code"
author: "T Goldberg and Carolyn Hicks" 
output: html_document
---

## Setup

Load Libraries:
```{r Load Libraries}
library(OncoDataSets)
library(dplyr)
library(ggplot2)
library(corrplot)
library(lattice)
library(caret)
library(kerndwd)
library(randomForest)
library(ranger)
library(ggfortify)
```
```{r Load Dataset}
df <- OncoDataSets::BreastCancerWI_df
```

## Dimensional Reduction and Feature Selection

### Entropy Analysis
(T Goldberg)

Define Entropy Function
```{r Define Entropy Function}
entropy <- function(X) {
  #create frequency table of X (count table devided by length of X) 
  freq <- table(X)/length(X)
  
  #make frequencies a vector
  vec <- as.data.frame(freq)[,2]
  
  #remove any zero counts if they occur because we will take log
  vec<-vec[vec>0]

  #entropy equation
  -sum(vec * log2(vec))
}
```

Define Information Gain
```{r Define Information Gain}

#I'm going to split the function the resource used into two to more easily test and view it. The first part returns dd_data

dd <-function(data, feature, parent, bins) {
 #clean out N/As
  data <- data[!is.na(data[,feature]),]
  
  #now break up our dataset into bins based on length
  data$cat<-cut(data[,feature], breaks=bins, labels=c(1:bins))
  
  #now we need to calculate the entropy of each feature (e)
  #we will need to know how many items fall in each bin (n)
  #to calculate frequency
  #we can also look at max and mins for each bin if we want
  
  dd_data <- data %>% group_by(cat) %>% summarise(e=entropy(get(parent)), n = length(get(parent)), min=min(get(feature)), max=max(get(feature))   
  )
  
  #now we need the probability for each feature value
  dd_data$p<-dd_data$n/nrow(data)
  
  return(dd_data)
}

#now the IG 
ig_num <- function(data, feature, parent, bins=5) {
  
  #entropy for our parent
  eP <- entropy(data[,parent])
  
  #call dd
  dd_data <- dd(data, feature, parent, bins)
  
  #and finally we can get our information gain
  IG <- eP - sum(dd_data$p*dd_data$e)
  
  return(IG)
}
```

Test Information Gain
```{r Test Information Gain: One Entry}
#test on one feature first
ig_num(df, "radius_mean", "diagnosis")

#look at dd_data
dd(df,"radius_mean", "diagnosis", 5)

#Based on results it looks like 5 bins may be too much for this category, as the feature entropy drops to zero for the last two bins. 
```

```{r Test Information Gain: Bin Numbers}
#Testing various bin numbers
ig_num(df, "radius_mean", "diagnosis", bins=3)

#look at dd_data
dd(df,"radius_mean", "diagnosis", 3)

#Interestingly, we still lose the last bin, and the information gain drops, so 5 bins is better!
```
```{r Test Information Gain: More Bins}
#We can look further at how bin number influences information gain for a single variable.

ig_num(df, "radius_mean", "diagnosis", bins=4)
ig_num(df, "radius_mean", "diagnosis", bins=7)
ig_num(df, "radius_mean", "diagnosis", bins=10)
ig_num(df, "radius_mean", "diagnosis", bins=15)

#As expected, as number of bins increases, the IG increases, but at a certain point there are fewer gains.
```

Final Information Gain Calculation
```{r Final Information Gain Calculation}
#Calculating IG for all of are variables with 5 bins

#define empty vectors
col_name<-character()
ig<-numeric()

features <- names(df)
length(features)

#for loop
for (i in 2:length(features)){
  col_name[i]<-features[i]
  ig[i]<-ig_num(df, names(df)[i], "diagnosis", bins=5)
}

ig_df<-arrange(data.frame(cbind(col_name, round(ig,2))),desc(V2))
ig_df
```

Information Gain Correlation
```{r Top IG Cross Correlation}
#These variables are highly likely to be cross correlated, so we should check correlations.

M = cor(select(df,concave_points_peak, c(ig_df$col_name[1:20])))
corrplot(M, method = 'color') # colorful number

#These variables are very highly correlated overall.
```

### Principal Component Analysis
(Carolyn Hicks)

Perform PCA
```{r Perform PCA}
pca <- prcomp(BreastCancerWI_df[, -1], center=TRUE, scale.=TRUE) #remove diagnosis column
print(pca)
summary(pca)
```

Visualize PCs
```{r Visualize PCs}
plot(pca,type="l") # lbow plot
autoplot(pca, data = BreastCancerWI_df, loadings = TRUE, loadings.label=TRUE, colour="diagnosis")
```

PC Selection and Training Dataset:

Unused Version
```{r PC Selection and Training Dataset (Unused Code), eval=FALSE}
#Note (TG): We ended up using a slightly different variable name for further analysis based on my training setup code, but this code is fully functional and equivalent.

top_pcs <- pca$x[,1:10]

library(dplyr)
df <- BreastCancerWI_df |> select(diagnosis)

top_pcs <- as.data.frame(top_pcs)
df <- as.data.frame(df)
training_df <-cbind(top_pcs, df)
training_df
```

Used Version
```{r PC Selection and Training Dataset (Used Code)}
#(T Goldberg)
data.pca <- data.frame(diagnosis = df$diagnosis, pca$x[,1:10])
```

Information Gain of PCs (T Goldberg)
```{r IG of PCs}
col_name2<-character()
ig2<-numeric()

features2 <- names(data.pca)
length(features2)

for (i in 2:length(features2)){
  col_name2[i]<-features2[i]
  ig2[i]<-ig_num(data.pca, names(data.pca)[i], "diagnosis", bins=5)
}

ig_df2<-arrange(data.frame(cbind(col_name2, round(ig2,2))),desc(V2))
ig_df2
#The top PC by far holds the most information.
```

Max Loadings
```{r Max Loadings}
loadings <- pca$rotation 
abs_loadings <- abs(loadings)
max_loadings <- apply(abs_loadings[, 1:10], 1, max) # first 10 PCs explain 95% of variance
sorted <- sort(max_loadings, decreasing = TRUE)

loadings_df <- as.data.frame(sorted)
loadings_df
```

Max Loading Correlation
```{r Max Loading Correlation}
#(TG) Correlation Plot for Max Loading Variables
corr_df <- select(BreastCancerWI_df, c(row.names(loadings_df)[1:10]))
cor_matrix <- cor(corr_df)
corrplot(cor_matrix, method='color')
```

## Model Selection and Initial Training
(T Goldberg)

Split Training and Validation Datasets
```{r Creation of Training and Validation Datasets}

set.seed(456)

#grabs indices of 70% of data with preserved class distribution for diagnosis
train.index <- createDataPartition(data.pca$diagnosis, p=0.70, list=FALSE)

#creates training dataset
training <- data.pca[train.index,]

#creates validation dataset
validation <- data.pca[-train.index,]
```

Training Harness
```{r Training Harness}

#We will need multiple seeds throughout process due to parallel processing.
#Caret package requires a seed per model tested, which is 30 models for 10-fold CV with k=3, plus a final seed
#Each seed should be an integer of length at least x digits with x based on the model type
#Difficult to ascertain the integer length, but 22 seemed to produce sufficient reproducibility
#The final integer should be a single digit

set.seed(456)

seedlist <- vector(mode = "list", length = 31)
for (i in 1:30) seedlist[[i]] <- sample.int(1000, 22)
seedlist[[31]] <- sample.int(1000, 1)

#as paper has done, we use repeated cross-validation with k=3 and 10 part splits and train for Accuracy

control <- trainControl(method = "repeatedcv", number=10, repeats = 3, seeds = seedlist)
metric <- "Accuracy"
```

Train LDA and RF Models with the Same Harness
```{r Train LDA and RF Models, cache=TRUE}
#note: seeds are reset within the harness
fit.lda2 <- train(diagnosis~., data=training, method="lda2", metric=metric, trControl=control)
fit.rf <- train(diagnosis~., data=training, method="rf", metric=metric, trControl=control)
```

Initial Model Comparison 
```{r Initial Model Comparison}
results <- resamples(list(lda=fit.lda2, rf=fit.rf))
summary(results)

dotplot(results)

print(fit.rf)
#Based on Accuracy and Kappa distributions, RF generally outperforms LDA
```

## Random Forest Optimization and Validation
(Carolyn Hicks)

### Untuned Model Performance 

Testing Untuned RF Model on Validation Set
```{r Testing Untuned Model on Validation Set}
#TG: added seed prior to running this for reproducibility
set.seed(123)

predictions_rf <- predict(fit.rf, newdata=validation)
confusionMatrix(predictions_rf, validation$diagnosis)

#Accuracy: 0.9647
```

### Optimizing Hyperparameters

Grid Search to Optimize mtry and Minimum Node Size

Define Grid:
```{r Define Grid}
# define grid values to test
grid <- expand.grid(
  mtry = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10),
  splitrule = "gini",
  min.node.size = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)
)
```

Unused Grid Seeds (T Goldberg)
```{r Grid Seeds, eval=FALSE}

#I slightly modified the grid harness by adding a seeds vector, but the single seed was reproducing results

set.seed(123)
seedlist2 <- vector(mode = "list", length = 6)
for (i in 1:5) seedlist2[[i]] <- sample.int(1000, 100)
seedlist2[[6]] <- sample.int(1000, 1)
```

Create Grid Training Control
```{r Create Grid Control}
ctrl <- trainControl(method = "cv", number = 5, search = "grid")
```

Grid Optimization Training
```{r Grid Optimization Training, cache=TRUE}

set.seed(123)
# train model based off of defined grid
grid_tuned_model <- train(
  diagnosis ~ .,
  data = training,
  method = "ranger", # use this method to incorporate min.node.size
  tuneGrid = grid,
  trControl = ctrl
)
```

Optimal Grid Parameters
```{r Optimal Parameters}
cat("Best Hyperparameters (Grid Search):\n")
print(grid_tuned_model$bestTune)

#Optimal mtry = 4 and min.node.size = 2
```

Optimizing ntrees with OOB
```{r Optimize ntrees with OOB, cache=TRUE}
#build model with optimized hyperparameters from grid search and 2000 decision trees

#TG: added seed prior to running this for reproducibility
set.seed(123)
rf4 <- randomForest(diagnosis~., data=training, 
                         ntree=2000, 
                         mtry=2,
                         nodesize=4)

#visualize results
plot(rf4)
#At a thousand trees it is relatively stable
```
Create Final Tuned Model
```{r Create Final Tuned Model, cache=TRUE}
#Use optimized parameters

#TG: added seed prior to running this for reproducibility
set.seed(123)

grid <- expand.grid(mtry=c(2))

final_tuned <- train(
  diagnosis~., 
  data=training, 
  method="rf", 
  trControl=control,
  tuneGrid=grid,
  ntree=1000,       
  min.node.size=4        
)
```

Final Model Accuracy
```{r, Final Model Accuracy}
results <- resamples(list(initial=fit.rf, tuned=final_tuned))
summary(results)

dotplot(results)
#Accuracy: 0.948 vs 0.946
#Kappa: 0.893 each
```
Final Model Validation
```{r, Final Model Validation, cache=TRUE}

#TG: added seed prior to running this for reproducibility
set.seed(123)

predictions <- predict(final_tuned, validation)
confusionMatrix(predictions, validation$diagnosis)

#Recall/sensitivity/true positive rate - 0.981
```
## Feature Importance
(Carolyn Hicks)

Feature Importance of PCs
```{r Feature Importance of PCs}
var_importance <- varImp(final_tuned)
print(var_importance)

ggplot(var_importance)
```
Loadings of PC1
```{r Loadings of PC1}
loadings <- pca$rotation[,1]
sorted <- sort(loadings, decreasing=FALSE)
print(sorted)
```

