---
title: "run_and_test_models"
output: html_document
date: "2026-04-22"
---

PCAs
```{r}
pca <- prcomp(BreastCancerWI_df[, -1], center=TRUE, scale.=TRUE) #removing diagnostics column
data.pca <- data.frame(diagnosis = df$diagnosis, pca$x[,1:10])
```

Source: https://machinelearningmastery.com/machine-learning-in-r-step-by-step/
Needed package 'caret'

Set up training and verification datasets (70:30)
```{r}
library(caret)
```

```{r}
#will need multiple seeds, easiest to set up vector of seeds from one seed
set.seed(456)

#grabs indices of 70% of data with preserved class distribution for diagnosis
train.index <- createDataPartition(data.pca$diagnosis, p=0.70, list=FALSE)

#creates training dataset
training <- data.pca[train.index,]

#creates validation dataset
validation <- data.pca[-train.index,]
```

Training Harness
(Setting up our training parameters)
```{r}

#We will need multiple seeds throughout process, can set up seed vector of length one per model
#we need one seed per model
#should be (model types x number of splits x number of repeats for repeated cross validation)
#paper does not do this!!!
#also makes it seem like they are doing 70/30 for training as well when its 9/10

set.seed(456)

seedlist <- vector(mode = "list", length = 31)
for (i in 1:30) seedlist[[i]] <- sample.int(1000, 22)
seedlist[[31]] <- sample.int(1000, 1)

#as paper has done, we use repeated cross-validation with k=3 and 10 part splits

control <- trainControl(method = "repeatedcv", number=10, repeats = 3, seeds = seedlist)
metric <- "Accuracy"
```

```{r}
library(randomForest)
library(caret)
library(ranger)
```

Train LDA and RF
```{r, echo=FALSE}
fit.lda2 <- train(diagnosis~., data=training, method="lda2", metric=metric, trControl=control)
fit.rf <- train(diagnosis~., data=training, method="rf", metric=metric, trControl=control)
```

Checks
```{r}
results <- resamples(list(lda=fit.lda2, rf=fit.rf))
summary(results)

dotplot(results)
```

Calculating MSE: https://www.geeksforgeeks.org/r-language/calculate-mse-for-random-forest-in-r-using-package-randomforest/ 
```{r}
validation$diagnosis <- as.numeric(validation$diagnosis)
```

```{r}
EnergyPred <- predict(fit.rf, validation)
print(mean((EnergyPred - validation$diagnosis)^2))
```
Visualization ideas: https://towardsdatascience.com/a-visual-guide-to-tuning-random-forest-hyperparameters/ 
Plotting MSE and most important features


Hyperparameter testing
Source: https://www.geeksforgeeks.org/machine-learning/hyperparameter-tuning-with-r/

Grid search
```{r}
grid <- expand.grid(
  mtry = c(1, 2, 3),
  splitrule = "gini",
  min.node.size = c(1, 5)
)

ctrl <- trainControl(method = "cv", number = 5, search = "grid")
set.seed(123)
grid_model <- train(
  diagnosis ~ .,
  data = training,
  method = "ranger",
  tuneGrid = grid,
  trControl = ctrl
)
cat("Best Hyperparameters (Grid Search):\n")
print(grid_model$bestTune)
```

Random search
```{r}
ctrl_random <- trainControl(method = "cv", number = 5, search = "random")

set.seed(123)
random_model <- train(
  diagnosis ~ .,
  data = training,
  method = "ranger",
  tuneLength = 10,   
  trControl = ctrl_random
)

cat("\nBest Hyperparameters (Random Search):\n")
print(random_model$bestTune)
```
how to take results from grid search and random search and incorporate that into running 

run validation model using grid search hyperparameters
```{r}
validation_grid_model <- randomForest(diagnosis~.,
                                      data=validation,
                                      mtry=3,
                                      nodesize=1) 
```

run validation model using random search hyperparameters
```{r}
validation_random_model <- randomForest(diagnosis~.,
                                       data=validation,
                                       mtry=9,
                                       nodesize=3)
```

visualize results