---
title: "stat437_project_tuning"
output: html_document
date: "2026-04-22"
---

libraries
```{r}
library(randomForest)
library(caret)
library(ranger)
library(ggplot2)
```

PCA
```{r}
pca <- prcomp(BreastCancerWI_df[, -1], center=TRUE, scale.=TRUE) #removing diagnostics column
data.pca <- data.frame(diagnosis = df$diagnosis, pca$x[,1:10])
```

Source: https://machinelearningmastery.com/machine-learning-in-r-step-by-step/
Needed package 'caret'

Set up training and verification datasets (70:30)
```{r}
#will need multiple seeds, easiest to set up vector of seeds from one seed
set.seed(456)

#grabs indices of 70% of data with preserved class distribution for diagnosis
train.index <- createDataPartition(data.pca$diagnosis, p=0.70, list=FALSE)

#creates training dataset
training <- data.pca[train.index,]

#creates validation dataset
validation <- data.pca[-train.index,]
```

Training Harness
(Setting up our training parameters)
```{r}

#We will need multiple seeds throughout process, can set up seed vector of length one per model
#we need one seed per model
#should be (model types x number of splits x number of repeats for repeated cross validation)
#paper does not do this!!!
#also makes it seem like they are doing 70/30 for training as well when its 9/10

set.seed(456)

seedlist <- vector(mode = "list", length = 31)
for (i in 1:30) seedlist[[i]] <- sample.int(1000, 22)
seedlist[[31]] <- sample.int(1000, 1)

#as paper has done, we use repeated cross-validation with k=3 and 10 part splits

control <- trainControl(method = "repeatedcv", number=10, repeats = 3, seeds = seedlist)
metric <- "Accuracy"
```

Train LDA and RF
```{r, echo=FALSE}
fit.lda2 <- train(diagnosis~., data=training, method="lda2", metric=metric, trControl=control)
fit.rf <- train(diagnosis~., data=training, method="rf", metric=metric, trControl=control)
```

Checks
```{r}
results <- resamples(list(lda=fit.lda2, rf=fit.rf))
summary(results)

dotplot(results)
```

Making predictions and confusion matrix using unseen data
```{r}
predictions_rf <- predict(fit.rf, newdata=validation)
confusionMatrix(predictions_rf, validation$diagnosis)
```
The paper looks at the following:
Precision - 0.981
Recall/sensitivity/true positive rate - 0.9626
False positive rate - 0.0317
Additionally, false negative rate - 0.0374

We will now use hyperparameter tuning to obtain smallest out-of-bag (OOB) estimate error and improve the predictive power of the dataset.

Starting with a grid search to find optimal mtry and min.node.size
Source: https://www.geeksforgeeks.org/machine-learning/hyperparameter-tuning-with-r/ 

```{r}
# define grid values to test
grid <- expand.grid(
  mtry = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10),
  splitrule = "gini",
  min.node.size = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)
)

ctrl <- trainControl(method = "cv", number = 5, search = "grid")
set.seed(123)

# train model based off of defined grid
grid_tuned_model <- train(
  diagnosis ~ .,
  data = training,
  method = "ranger", # use this method to incorporate min.node.size
  tuneGrid = grid,
  trControl = ctrl
)

# look at optimal hyperparameters
cat("Best Hyperparameters (Grid Search):\n")
print(grid_tuned_model$bestTune)
```

Next, we will determine the optimal ntrees value to minimize OOB error and computational cost. Elbow plot shows point of diminishing returns, where adding more trees will not increase model performance but will heighten computational cost. 

```{r}
#build model with optimized hyperparameters from grid search and 2000 decision trees
rf4 <- randomForest(diagnosis~., data=training, 
                         ntree=2000, 
                         mtry=2,
                         nodesize=4)

#visualize results
plot(rf4)
```
1000 trees looks to be where OOB stabilizes. 

Using the optimized mtry, minimum nodes, and ntree values, build the final tuned model. 
```{r}
grid <- expand.grid(mtry=c(2))

final_tuned <- train(
  diagnosis~., 
  data=training, 
  method="rf", 
  trControl=control,
  tuneGrid=grid,
  ntree=1000,       
  min.node.size=4        
)
```

Visualize accuracy metrics of the models
```{r}
results <- resamples(list(initial=fit.rf, tuned=final_tuned))
summary(results)

dotplot(results)
```
Accuracy and kappa metrics:
Our tuned model has an accuracy of 0.9474 which is slightly higher than the initial model which has a mean accuracy of 0.9457. Kappa increased from 0.8833 to 0.8873. Although this seems like a small change in accuracy, the train function used to build the model already builds a highly accurate model using a combination of cross-validation and automatic hyperparameter tuning.

Apply tuned model to validation data
```{r}
predictions <- predict(final_tuned, validation)
confusionMatrix(predictions, validation$diagnosis)
```
The paper looks at the following:
Precision - 0.972
Recall/sensitivity/true positive rate - 0.981
False positive rate - 0.0476
Additionally, false negative rate - 0.0187

Look at which variables are most important for random forest model, looking at mean decrease in accuracy. While the paper looks at mean decrease in gini index for variable importance, we trained our models using the train() function (to incorporate training controls) so importance() function can't be used. 
```{r}
var_importance <- varImp(final_tuned)
print(var_importance)

ggplot(var_importance)
```

Look at loadings of PC1
```{r}
loadings <- pca$rotation[,1]
sorted <- sort(loadings, decreasing=FALSE)
print(sorted)
```


