---
title: "Final Project Code"
author: "T Goldberg and Carolyn Hicks" 
output: html_document
---

## Setup

Load Libraries:
```{r Load Libraries}
library(OncoDataSets)
library(dplyr)
library(ggplot2)
library(corrplot)
library(lattice)
library(caret)
library(kerndwd)
library(randomForest)
library(ranger)
library(ggfortify)
```
```{r Load Dataset}
df <- OncoDataSets::BreastCancerWI_df
```

## Dimensional Reduction and Feature Selection

### Entropy Analysis
(T Goldberg)

Define Entropy Function
```{r Define Entropy Function}
entropy <- function(X) {
  #create frequency table of X (count table devided by length of X) 
  freq <- table(X)/length(X)
  
  #make frequencies a vector
  vec <- as.data.frame(freq)[,2]
  
  #remove any zero counts if they occur because we will take log
  vec<-vec[vec>0]

  #entropy equation
  -sum(vec * log2(vec))
}
```

Define Information Gain
```{r Define Information Gain}

#I'm going to split the function the resource used into two to more easily test and view it. The first part returns dd_data

dd <-function(data, feature, parent, bins) {
 #clean out N/As
  data <- data[!is.na(data[,feature]),]
  
  #now break up our dataset into bins based on length
  data$cat<-cut(data[,feature], breaks=bins, labels=c(1:bins))
  
  #now we need to calculate the entropy of each feature (e)
  #we will need to know how many items fall in each bin (n)
  #to calculate frequency
  #we can also look at max and mins for each bin if we want
  
  dd_data <- data %>% group_by(cat) %>% summarise(e=entropy(get(parent)), n = length(get(parent)), min=min(get(feature)), max=max(get(feature))   
  )
  
  #now we need the probability for each feature value
  dd_data$p<-dd_data$n/nrow(data)
  
  return(dd_data)
}

#now the IG 
ig_num <- function(data, feature, parent, bins=5) {
  
  #entropy for our parent
  eP <- entropy(data[,parent])
  
  #call dd
  dd_data <- dd(data, feature, parent, bins)
  
  #and finally we can get our information gain
  IG <- eP - sum(dd_data$p*dd_data$e)
  
  return(IG)
}
```

Test Information Gain
```{r Test Information Gain: One Entry, eval=FALSE}
#test on one feature first
ig_num(df, "radius_mean", "diagnosis")

#look at dd_data
dd(df,"radius_mean", "diagnosis", 5)

#Based on results it looks like 5 bins may be too much for this category, as the feature entropy drops to zero for the last two bins. 
```

```{r Test Information Gain: Bin Numbers, eval=FALSE}
#Testing various bin numbers
ig_num(df, "radius_mean", "diagnosis", bins=3)

#look at dd_data
dd(df,"radius_mean", "diagnosis", 3)

#Interestingly, we still lose the last bin, and the information gain drops, so 5 bins is better!
```
```{r Test Information Gain: More Bins, eval=FALSE}
#We can look further at how bin number influences information gain for a single variable.

ig_num(df, "radius_mean", "diagnosis", bins=4)
ig_num(df, "radius_mean", "diagnosis", bins=7)
ig_num(df, "radius_mean", "diagnosis", bins=10)
ig_num(df, "radius_mean", "diagnosis", bins=15)

#As expected, as number of bins increases, the IG increases, but at a certain point there are fewer gains.
```

Final Information Gain Calculation
```{r Final Information Gain Calculation}
#Calculating IG for all of are variables with 5 bins

#define empty vectors
col_name<-character()
ig<-numeric()

features <- names(df)
length(features)

#for loop
for (i in 2:length(features)){
  col_name[i]<-features[i]
  ig[i]<-ig_num(df, names(df)[i], "diagnosis", bins=5)
}

ig_df<-arrange(data.frame(cbind(col_name, round(ig,2))),desc(V2))
ig_df
```

Information Gain Correlation
```{r Top IG Cross Correlation}
#These variables are highly likely to be cross correlated, so we should check correlations.

M = cor(select(df,concave_points_peak, c(ig_df$col_name[1:20])))
corrplot(M, method = 'color') # colorful number

#These variables are very highly correlated overall.
```

### Principal Component Analysis
(Carolyn Hicks)

Perform PCA
```{r Perform PCA}
pca <- prcomp(BreastCancerWI_df[, -1], center=TRUE, scale.=TRUE) #remove diagnosis column
print(pca)
summary(pca)
```

Visualize PCs
```{r Visualize PCs}
plot(pca,type="l") # lbow plot
autoplot(pca, data = BreastCancerWI_df, loadings = TRUE, loadings.label=TRUE, colour="diagnosis")
```

PC Selection and Training Dataset:

(Unused Version)
```{r PC Selection and Training Dataset (Unused Code), eval=FALSE}
#Note (TG): We ended up using a slightly different variable name for further analysis based on my training setup code, but this code is fully functional and equivalent.

top_pcs <- pca$x[,1:10]

library(dplyr)
df <- BreastCancerWI_df |> select(diagnosis)

top_pcs <- as.data.frame(top_pcs)
df <- as.data.frame(df)
training_df <-cbind(top_pcs, df)
training_df
```

(Used Version)
```{r PC Selection and Training Dataset (Used Code)}
#(T Goldberg)
data.pca <- data.frame(diagnosis = df$diagnosis, pca$x[,1:10])
```

Information Gain of PCs (T Goldberg)
```{r IG of PCs}
col_name2<-character()
ig2<-numeric()

features2 <- names(data.pca)
length(features2)

for (i in 2:length(features2)){
  col_name2[i]<-features2[i]
  ig2[i]<-ig_num(data.pca, names(data.pca)[i], "diagnosis", bins=5)
}

ig_df2<-arrange(data.frame(cbind(col_name2, round(ig2,2))),desc(V2))
ig_df2
#The top PC by far holds the most information.
```

## Model Selection and Initial Training

## Random Forest Optimization and Validation

## Feature Importance
